On the properties of discs around accreting brown dwarfs

### 2010-06-11 BD_disc/paper.tex
mods by tim - th
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2296 25464f21-6099-402c-b30a-640f1d40d38b

  We present a grid of simulated accreting brown dwarf systems with
  (and without) circumstellar discs. These simulations include solving
  for vertical hydrostatic and radiative equilibirum
  self-consistently, and a sophisticated implementation of dust
  sublimation. We have simulated observations of the spectral energy
  distributions and using several broadband photometric systems. The
  resulting magnitudes and spectra are publicly available.

  Analysis of the disc structures and simulated observations reveal a
  natural dichotomy in accretion rates, with \logmdot $>-$9 and $\leq
  -$9 classed as extreme and typical accretors respectively. The
  photospheric emissions of the extreme accretors are shown to be
  heavily veiled, and significant sublimation of the dust disc is
  apparent in these systems.

  Derivation of ages and masses from our simulated photometry, using
  isochrones, is demostrated to be unreliable even for typical
  accretors. Current brown dwarf disc candidate selection criteria
  have been shown to be largely reliable when applied to our model
  grid, for both accretion regimes. However, we suggest improved
  selection criteria in several colour indices, derived from our grid.

  We also show that as accretion rates increase, due to photometric
  scatter, brown dwarf disc systems are less likely to be correctly
  identified. This suggests that, within our grid, systems with higher
  accretion rates would be preferentially lost during brown dwarf
  target selection. Therefore, we suggest that observations used to
  assert a $\dot{M}\propto M_*^{\sim 2}$ relationship may contain an
  intrinsic selection bias.

@@

The truncated inner-edge of the circumstellar disc receives direct
photospheric radiation and is strongly heated. This leads to an
increased scaleheight, or `puffing up' of the inner disc rim, and
subsequent shadowing of the region immediately beyond it
\citep{dullemond_2001}. The density-dependent nature of the dust
sublimation temperature \citep{pollack_1994} may also play a role in
shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) sytems to that of the
  CTTS-like phase we are considering here.}.

@@

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^{\sim 2}_{*}$. Since in
the canonical picture the accretion rate is driven by the disc
viscosity, and should be indepedent of the mass of the central object,
this correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}: At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photopsheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may be
a population of low-accretion-rate CTTS which are current missing from
the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed \citep{muzerolle_2003,natta_2004,natta_2006}. (Cooling rate
arguments can be invoked to explain the presence of such high
temperatures). Recently some continuum measurements of accreting BDs
have been conducted \cite{herczeg_2009}, and these broadly support the
H$\alpha$ rates.

@@

In this section we briefly explain the key elements of the radiative
transfer code used to model the BDD systems.

We have used the {\sc torus} radiative transfer code which is
described in \cite{harries_2000}, including the subsequent refinements
i.e.  addition of Adaptive-Mesh-Refinement (AMR) introduced in
\cite{harries_2004}. {\sc torus} uses the method of \cite{lucy_1999}
solve radiative equilibrium. The simulation also self-consistently
solves the equation of vertical hydrostatic equilibrium and dust
sublimation for the disc \citep[described in][]{tannirkulam_2007}.

@@

We have adopted a similar dust model to \cite{wood_2002} 
with the size distribution of dust particle given by,
\begin{equation}
\label{particle_dist}
n(a)da=C_ia^{-q}\times exp^{[-(a/a_c)^p]}da
\end{equation},
where $n(a)da$ is the number of particles of size $a$ (within the
increment $da$), $a_{\rm c}$ is the characteristic particle size, with
$p$ and $q$ simply used to control the shape of the
distribution. $C_i$ controls the relative abundance of each
constituent species ($i$) in the dust. \cite{wood_2002}
found that simulated SEDs fit observed SEDs better using this adjusted
size distribution for the dust particles, as opposed to a simple power
law. The best fitting values found in \cite{wood_2002} were
$q=$3.0, $p=$0.6 with $a_{\rm c}=$50\,$\mu$m, also with an associated
maximum and minimum grain size of $a_{\rm min}=$5 nm and $a_{\rm
  max}=$1mm. A slightly steeper power law dependence, with $q=$3.5
(corresponding to the canonical MRN distribution) is more widely
adopted
\citep{bouy_2008,morrow_2008,pascucci_2008}.
Here we have adopted the values of \cite{wood_2002}
except for the parameter $q$ where we have used $q=$3.5.
\cite{wood_2002} calculated the $C_i$ values for amorphous
carbon and silicon by requiring the dust to deplete a solar abundance
of either component completely \citep[using abundances
  from][]{anders_1989,noels_1993}. We have set
$C_i=$1 and adjusted the species using a grain fractional abundance
\cite[an equivalent process to that of][]{wood_2002},
however we have adjusted these grain fractions using the updated solar
abundances of \cite{asplund_2006}. The resulting difference in
opacity between grain fractions matching the work of
\cite{wood_2002} and the new grain fractions is only a
slight enhancement of the silicate feature (due to the relative
abundance of silicon increasing) which has little effect on the
resulting SEDs. Figure \ref{albedo} shows the resulting albedo, and
scattering and absorption opacities, for our dust population, with a
the vertical dashed line showing 10\,$\mu$m.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range
\citep{dullemond_2001}.

@@

Many observational studies use non-spectroscopic data to derive the
pertinent parameters. Therefore in order to examine the practical
effects in the `observational plane' we have used the SEDs to produce
broadband photometric magnitudes, and subsequently colours. Broadband
magnitudes were also derived in a large range of other filter sets not
used explicitly in the analysis within this paper. These magnitudes
are available online
\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore disc}
and are briefly discussed in Appendix \ref{website}. In addition,
monochromatic fluxes have been derived for all filters, and again are
available online and discussed in Appendix \ref{website}.

@@

In some cases the periodic variability is irregular and assumed to
come from active accretion hot spots on the BD surface \citep[see][for
a discussion of variability causes]{bouvier_1995,herbst_2007},
indicative of active accretion. All the studies mentioned infer a disc
locking mechanism. Furthermore, \cite{scholz_2004} and
\cite{scholz_2005} find evidence for a mass$\propto$period
relationship extending into the BD regime. Additionally,
\cite{joergens_2003} propose a shorter lifetime of $\approx$5 Myrs for
BD discs, inferred from a shorter derived disc locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs, for
BD discs. The data on BD rotation rates, disc presence and disc
locking are summarised and discussed in \cite{herbst_2007}.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
  TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was perhaps first suggested by
\cite{muzerolle_2003} using various accretion diagnostics. Later,
\citep{muzerolle_2005} derived a relationship of approximately
$\dot{M}\propto M_*^{\sim 2}$. Further evidence was put forward by
\cite{natta_2004}, where accretion rates as low as 5$\times$
10$^{-12}M_{\odot} yr^{-1}$ were found for BD stars. More recently,
even lower accretion rates of $\approx$10$^{-13}M_{\odot} yr^{-1}$,
have been derived for BDs by \cite{herczeg_2009}.  Further support for
a dependence of accretion rate on stellar mass was apparent in the
significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for pre-MS stars in $\rho$ Ophiuchi and compared
these results to stars in Taurus. They found that the accretion rate
scales with central object mass into the BD regime, although with
significant scatter.

As the relationship $\dot{M}\propto M_*^{\sim 2}$ predicts lower accretion
rates for BD mass objects it is essential that we model systems at
higher accretion rates, which may have been missed in current
observational studies. Therefore, we have adopted accretion rates of
\logmdot = $-6$, $-7$, $-8$, $-9$, $-10$, $-11$ \& $-12$.

@@

Previously studies modeling BD discs have adopted a range of disc mass
fractions, for instance \citep{walker_2004} use 0.1, 0.01 and
0.001$M_*$. \cite{wood_2002} fitted observed spectra with modeled SEDs
to derive a disc mass of 0.003$M_*$, for HH 30 IRS. Subsequent
derivations of disc masses have converged to within an order of
magnitude, with the following specific results: 0.03$M_*$
\citep[$\rho$ Ophiuchi,][]{natta_2002}, 0.055$M_*$ \citep[GM
Aurigae,][]{rice_2003}, 0.03$M_*$ \citep[GY 5, GY 11, and GY
310,][]{mohanty_2004} and 0.022$M_*$\citep[2MASS
J04442713+2512164,][]{bouy_2008}. As the derived disc masses all have
a similar order of magnitude we have adopted $M_{\rm disc}\approx
$0.01$M_*$. As changes in disc masses are expected to change the
resulting SED less than perhaps, accretion rate for example, we have
not varied the disc mass for this study. The results of simulations
varying this parameter will be published in a future paper.

@@

A further complexity arises when one considers that the density of
material in the disc falls with increasing radius from the star
($\rho(r) \propto r^{\alpha}$), and that the dust sublimation
temperature is density dependent \citep{pollack_1994}.  Therefore, for
systems where the inner edge has been eroded significantly from the
co-rotation radius the temperature of the inner edge will fall
systematically as the inner dust disc radius increases.

For further analysis, and throughout the rest of this paper, we will
seperate the systems into two groups. These groups are those systems
with \logmdot$\leq-9$ and \logmdot$\geq-9$, classed as typical and
extreme accretors. This distinction is based on the point at which the
systems undergo significant sublimation ($\Delta R_{\rm
  inner}>1R_*$). In order to examine the effect of dust sublimation on
our models we have measured the radius of the inner edge of the dust
disc by integrating from the centre along the midplane until an
optical depth of 2/3 (at 5500\AA) is reached. Figure \ref{inner_temp}
shows the radial density distribution of the disc.  In this case the
the final inner edge radius and therefore temperature is no longer
strongly correlated with the rotation rate. Figure \ref{inner_temp}
shows the final inner edge location ($R_{\rm inner}$, $R_*$) against
the temperature of the inner wall ($T_{\rm inner}$, K). The left panel
shows the systems designated as typical accretors and the right panel
those with extreme accretion rates.  For both panels the systems with
rotation periods of 0.5 and 5 days are plotted in red and blue
respectively. Those systems where the change in inner radius was
greater than 1$R_*$, and therefore classes as significantly
sublimated, are plotted as crosses.

@@

Disentangling the accretion and stellar photospheric flux, in order to
derive accretion rates becomes difficult as $T_{\rm acc} \to T_{\rm
  eff}$, which for an accretion rate of \logmdot = $-9$, occurs at a
fractional coverages of 20 and 10\%, for rotation periods of 5 and 0.5
days (with $M_*=0.04M_{\odot}$ and an age of 1 Myr). Additionally,
higher accretion rates can produce enough flux to `veil' the
underlying photospheric features, and significantly change the peak
flux levels. Heavy veiling of atomic bands has been extensively
observed in CTTS systems \citep[e.g][]{kenyon_1995}.

@@

The left panels show accretion streams with an areal coverage of 10\%
and the left panels 1\%. As the areal coverage reduces the effective
temperature of the accretion hot spot increases, resulting in an
increase in accretion flux, and resulting shift to bluer wavelengths
of the peak flux. This can be seen clearly by comparing the left and
right panels of Figure \ref{acc_dom} As the rotation rate increases
the co-rotation radius also increases, as shown in Equation
\ref{inner_eq}, $R_{\rm inner}\propto \tau^{\frac{2}{3}}$. Therefore,
as Equations \ref{Lacc} and \ref{Tacc}, show, $L_{\rm acc}\propto 1-
\frac{R_*}{R_{\rm inner}}$ and $T_{\rm acc}\propto (\frac{L_{\rm
    acc}}{A})^{\frac{1}{4}}$. Therefore, the temperature of the
accretion hot spot increases as the rotational period increases, due
to the increase in potential energy lost by the mass accreted. This
can be seen in Figure by comparing the left and right panels of Figure
\ref{acc_dom}. Perhaps the most important, albeit qualitative, result
shown is Figure \ref{acc_dom} is an insight into the accretion rate at
which the accretion blackbody flux dominates over the photospheric
flux. Figure \ref{acc_dom} shows that as the accretion rate raises
above \logmdot = $-9$ for systems with 1 or 10\% areal coverage, the
accretion flux dominates the emergent SED at both
periods. Effectively, the spectroscopic features of the photosphere
are veiled by the additional continuum accretion flux. Therefore for
reasonable coverages (1--10\%) and rotational periods (0.5--5 days)
the photospheric flux is effectively veiled by accretion flux for
accretion rates \logmdot$>-9$. It is also clear from Figure
\ref{acc_dom} that the magnitude becomes brighter and the colour bluer
(in terms of optical photometry) with increasing accretion rates as
expected \citep{gullbring_1998}. The impact on the photometry of BDD
systems is discussed in the next Section for individual stars and in
Section \ref{isochrones} for populations.

@@

As shown in Figures \ref{flare_-12_183} and \ref{flare_-7_193} and
discussed in Section \ref{disc_flaring} BDD systems under vertical
hydrostatic equilibrium have highly flared discs. As discussed in
\cite{walker_2004} this increased flaring (when compared to CTTS
stars) leads to occultation at the star at lower inclination angles
and, therefore, significant changes to the SED. Increases in the
inclination angle for these systems quickly lead to a significant
proportion of the stellar flux being intercepted and reprocessed by
the highly flared disc. This reprocessing will lead to a change in the
flux levels at the shorter, bluer, wavelengths as more stellar flux is
intercepted by the disc. It will also lead to significant changes in
the flux reaching the observer from the inner and outer regions of the
disc as the system approaches edge on. Also, as discussed in Section
\ref{disc_struct} the addition of dust sublimation leads to a change
in the shape of the inner edge, where the temperature is
sufficient. Figures \ref{no_sub_183_a} to \ref{sub_179_a}
\ref{sub_181_a} show a range in inner edge shapes, from flat walls
through concave to convex curves, caused by the radial density profile
of the disc and the dependence of the sublimation temperature on
density. This change in shape, as noted for Herbig Ae stars by
\cite{tannirkulam_2007}, will lead to changes in the characteristics
of the SED, or derived IR excess, with inclination angle
\citep{tannirkulam_2007}, for close to face on viewing angles.

@@

Figure \ref{inner_inc} shows the SEDs for the viewing angles of 0, 27,
39, and 48 $^{\circ}$, close to face on. The top and bottom panels of
Figure \ref{inner_inc} show SEDs for models with a curved, as
exemplified by Figure \ref{flare_-7_193}, and a flat inner wall, as
shown in Figure \ref{flare_-12_183} respectively. The specific models
chosen have been selected to minimise flux differences between the two
SEDs but have distinct inner edges. The top panel has a mass of
0.04$M_{\odot}$, \logmdot = -8, areal coverage of 1\%, a rotational
period of 5 days and an age of 10 Myrs. The bottom panel has a the
same model but with rotation period of 0.5 days and an age of 1
Myrs. Comparing the top and bottom panel show that flux levels for the
higher accretion rate, with the larger and curved inner edge, vary
less distinctly with inclination angle than the systems with a
vertical wall. The difference in inclination dependence of the fluxes
between the curved and vertical inner walls is marginal. This is due
to differences in the stellar flux levels and disc
structures. Essentially we are not comparing like for like systems,
ideally we would replace the curved inner edge with a vertical wall in
an eample system and recalculate the temperature structure and SED.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/inner_inc.ps}
  \caption{Figure showing SEDs of systems with a curved and flat inner
    wall as the top and bottom panels respectively. The lines show the
    SEDs at viewing angles of 0, 27, 39, and 48 $^{\circ}$, close to
    face-on.\label{inner_inc}}
\end{figure*}

@@

As stated throughout this paper the data presented are available from
a web page
\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore disc}.

>>>>

  We present a grid of models of accreting brown dwarf systems with
  circumstellar discs. The calculations involve a self-consistent
  solution of both vertical hydrostatic and radiative equilibirum
  along with a sophisticated treatment of dust sublimation. We have
  simulated observations of the spectral energy distributions and
  several broadband photometric systems. Analysis of the disc
  structures and simulated observations reveal a natural dichotomy in
  accretion rates, with \logmdot $>-$9 and $\leq -$9 classed as
  extreme and typical accretors respectively.  Derivation of ages and
  masses from our simulated photometry using isochrones is demostrated
  to be unreliable even for typical accretors. Although current brown
  dwarf disc candidate selection criteria have been shown to be
  largely reliable when applied to our model grid we suggest improved
  selection criteria in several colour indices. We show that as
  accretion rates increase brown dwarf disc systems are less likely to
  be correctly identified. This suggests that, within our grid,
  systems with higher accretion rates would be preferentially lost
  during brown dwarf target selection. We suggest that observations
  used to assert a $\dot{M}\propto M_*^2$ relationship may
  contain an intrinsic selection bias.

@@

@@

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be indepedent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}: At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photopsheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may
well be a population of low-accretion-rate CTTS which are current
missing from the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed \citep{muzerolle_2003,natta_2004,natta_2006}. Although
cooling rate arguments can be invoked to explain the presence of such
high temperatures, the line ratios of Pa$\beta$ and Br$\gamma$
indicate that the hydrodgen emission may not arise in the funnel flows
\citep{gatti_2006}. However, recently some continuum measurements of
accreting BDs have been conducted \citep{herczeg_2009}, and these
broadly support the H$\alpha$ rates.

@@

We have used the {\sc torus} radiative transfer code which was
originally described by \cite{harries_2000} and was subsequently
updated by \cite{harries_2004} and \cite{kurosawa_2004}. {\sc torus}
  uses the method of \cite{lucy_1999} solve radiative equilibrium on
  an AMR mesh. The code also self-consistently solves the
  equation of vertical hydrostatic equilibrium and dust sublimation
  for the disc \citep[described in][]{tannirkulam_2007}. The {\sc
    torus} code has been extensively benchmarked \citep{pinte_2009}.

@@

We have adopted a similar dust model to \cite{wood_2002} 
with the size distribution of dust particle given by
\begin{equation}
\label{particle_dist}
n(a)da=C_ia^{-q}\times exp^{[-(a/a_c)^p]}da
\end{equation}
where $n(a)da$ is the number of particles of size $a$ (within the
increment $da$), $a_{\rm c}$ is the characteristic particle size, with
$p$ and $q$ simply used to control the shape of the
distribution. $C_i$ controls the relative abundance of each
constituent species ($i$) in the dust. \cite{wood_2002}
found that simulated SEDs fit observed SEDs better using this adjusted
size distribution for the dust particles, as opposed to a simple power
law. The best fitting values found in \cite{wood_2002} were
$q=$3.0, $p=$0.6 with $a_{\rm c}=$50\,$\mu$m, also with an associated
maximum and minimum grain size of $a_{\rm min}=$5 nm and $a_{\rm
  max}=$1mm. A slightly steeper power law dependence, with $q=$3.5
(corresponding to the canonical MRN distribution) is more widely
adopted
\citep{bouy_2008,morrow_2008,pascucci_2008}.
Here we have adopted the values of \cite{wood_2002}
except for the parameter $q$ where we have used $q=$3.5.
\cite{wood_2002} calculated the $C_i$ values for amorphous
carbon and silicon by requiring the dust to deplete a solar abundance
of either component completely \citep[using abundances
  from][]{anders_1989,noels_1993}. We have set
$C_i=$1 and adjusted the species using a grain fractional abundance
\cite[an equivalent process to that of][]{wood_2002},
however we have adjusted these grain fractions using the updated solar
abundances of \cite{asplund_2006}. The resulting difference in
opacity between grain fractions matching the work of
\cite{wood_2002} and the new grain fractions is only a
slight enhancement of the silicate feature (due to the relative
abundance of silicon increasing) which has little effect on the
resulting SEDs. Figure \ref{albedo} shows the resulting albedo, and
scattering and absorption opacities, for our dust population, with a
the vertical dashed line showing 10\,$\mu$m.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range \citep{dullemond_2001}. The
circumstellar disc receives direct photospheric radiation and is
strongly heated. This leads to an increased scaleheight, or `puffing
up' of the inner disc rim, and subsequent shadowing of the region
immediately beyond it \citep{dullemond_2001}. The density-dependent
nature of the dust sublimation temperature \citep{pollack_1994} may
also play a role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) sytems to that of the
  CTTS-like phase we are considering here.}.

@@

Many observational studies use non-spectroscopic data to derive the
pertinent parameters. Therefore in order to examine the practical
effects in the `observational plane' we have used the SEDs to produce
broadband photometric magnitudes, and subsequently colours. Broadband
magnitudes were also derived in a large range of other filter sets not
used explicitly in the analysis within this paper. These magnitudes
are available online\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore disc}
and are briefly discussed in Appendix \ref{website}. In addition,
monochromatic fluxes have been derived for all filters, and again are
available online and discussed in Appendix \ref{website}.

@@

In some cases the periodic variability is irregular and assumed to
come from active accretion hot spots on the BD surface \citep[see][for
a discussion of variability causes]{bouvier_1995,herbst_2007},
indicative of active accretion. All the studies mentioned infer a disc
locking mechanism. Furthermore, \cite{scholz_2004} and
\cite{scholz_2005} find evidence for a mass $\propto$ period
relationship extending into the BD regime. Additionally,
\cite{joergens_2003} propose a shorter lifetime of $\approx$5 Myrs for
BD discs, inferred from a shorter derived disc-locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs, for
BD discs. The data on BD rotation rates, disc presence and disc
locking are summarised and discussed in \cite{herbst_2007}.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
  TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was perhaps first suggested by
\cite{muzerolle_2003} using various accretion diagnostics. Later,
\citep{muzerolle_2005} derived a relationship of approximately
$\dot{M}\propto M_*^2$. Further evidence was put forward by
\cite{natta_2004}, where accretion rates as low as 5$\times$
10$^{-12} {\rm M}_{\odot} {\rm yr}^{-1}$ were found for BD stars. More recently,
even lower accretion rates of $\approx$10$^{-13} {\rm M}_{\odot} {\rm yr}^{-1}$,
have been derived for BDs by \cite{herczeg_2009}.  Further support for
a dependence of accretion rate on stellar mass was apparent in the
significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for pre-MS stars in $\rho$ Ophiuchi and compared
these results to stars in Taurus. They found that the accretion rate
scales with central object mass into the BD regime, although with
significant scatter.

As the relationship $\dot{M}\propto M_*^2$ predicts lower accretion
rates for BD mass objects it is essential that we model systems at
higher accretion rates, which may have been missed in current
observational studies. Therefore, we have adopted accretion rates of
\logmdot = $-6$, $-7$, $-8$, $-9$, $-10$, $-11$ \& $-12$.

@@

Previously studies modeling BD discs have adopted a range of disc mass
fractions, for instance \citep{walker_2004} use 0.1, 0.01 and
0.001$M_*$. \cite{wood_2002} fitted observed spectra with modeled SEDs
to derive a disc mass of 0.003$M_*$  for HH~30 IRS. Subsequent
derivations of disc masses have converged to within an order of
magnitude, with the following specific results: 0.03$M_*$
\citep[$\rho$ Ophiuchi,][]{natta_2002}, 0.055$M_*$ \citep[GM
Aurigae,][]{rice_2003}, 0.03$M_*$ \citep[GY 5, GY 11, and GY
310,][]{mohanty_2004} and 0.022$M_*$\citep[2MASS
J04442713+2512164,][]{bouy_2008}. As the derived disc masses all have
a similar order of magnitude we have adopted $M_{\rm disc}\approx
$0.01$M_*$. As changes in disc masses are expected to change the
resulting SED less than perhaps, accretion rate for example, we have
not varied the disc mass for this study. The results of simulations
varying this parameter will be published in a future paper.

@@

A further complexity arises when one considers that the density of
material in the disc falls with increasing radius from the star
($\rho(r) \propto r^{-\alpha}$), and that the dust sublimation
temperature is density dependent \citep{pollack_1994}.  Therefore, for
systems where the inner edge has been eroded significantly from the
co-rotation radius the temperature of the inner edge will fall
systematically as the inner dust disc radius increases.

For further analysis, and throughout the rest of this paper, we will
seperate the systems into two groups. These groups are those systems
with \logmdot$\leq-9$ and \logmdot$\geq-9$, classed as typical and
extreme accretors. This distinction is based on the point at which the
systems undergo both significant sublimation ($\Delta R_{\rm
  inner}>1R_*$) and have an accretion luminosity that is comparible to
the photospheric luminosity (see Section~\ref{acc_dominance}). In
order to examine the effect of dust sublimation on our models we have
measured the radius of the inner edge of the dust disc by integrating
from the centre along the midplane until an optical depth of 2/3 (at
5500\AA) is reached. Figure \ref{inner_temp} shows the radial density
distribution of the disc.  In this case the the final inner edge
radius and therefore temperature is no longer strongly correlated with
the rotation rate. Figure \ref{inner_temp} shows the final inner edge
location ($R_{\rm inner}$, $R_*$) against the temperature of the inner
wall ($T_{\rm inner}$, K). The left panel shows the systems designated
as typical accretors and the right panel those with extreme accretion
rates.  For both panels the systems with rotation periods of 0.5 and 5
days are plotted in red and blue respectively. Those systems where the
change in inner radius was greater than 1$R_*$, and therefore classes
as significantly sublimated, are plotted as crosses.

@@

Disentangling the accretion and stellar photospheric flux, in order to
derive accretion rates becomes difficult as $T_{\rm acc} \to T_{\rm
  eff}$, which for an accretion rate of \logmdot = $-9$, occurs at a
fractional coverages of 20 and 10\%, for rotation periods of 5 and 0.5
days (with $M_*=0.04M_{\odot}$ and an age of 1 Myr). Additionally,
higher accretion rates can produce enough flux to `veil' the
underlying photospheric features, and significantly change the peak
flux levels. Heavy veiling of atomic lines has been extensively
observed in CTTS systems \citep[e.g][]{kenyon_1995}.

@@

The left panels show accretion streams with an areal coverage of 10\%
and the left panels 1\%. As the areal coverage reduces the effective
temperature of the accretion hot spot increases, resulting in an
increase in accretion flux, and resulting shift to bluer wavelengths
of the peak flux. This can be seen clearly by comparing the left and
right panels of Figure \ref{acc_dom} As the rotation rate increases
the co-rotation radius also increases, as shown in Equation
\ref{inner_eq}, $R_{\rm inner}\propto \tau^{\frac{2}{3}}$. Therefore,
as Equations \ref{Lacc} and \ref{Tacc}, show, $L_{\rm acc}\propto 1-
\frac{R_*}{R_{\rm inner}}$ and $T_{\rm acc}\propto (L_{\rm
    acc}/A)^{\frac{1}{4}}$. Therefore, the temperature of the
accretion hot spot increases as the rotational period increases, due
to the increase in potential energy lost by the mass accreted. This
can be seen in Figure by comparing the left and right panels of Figure
\ref{acc_dom}. Perhaps the most important, albeit qualitative, result
shown is Figure \ref{acc_dom} is an insight into the accretion rate at
which the accretion blackbody flux dominates over the photospheric
flux. Figure \ref{acc_dom} shows that as the accretion rate raises
above \logmdot = $-9$ for systems with 1 or 10\% areal coverage, the
accretion flux dominates the emergent SED at both
periods. Effectively, the spectroscopic features of the photosphere
are veiled by the additional continuum accretion flux. Therefore for
reasonable coverages (1--10\%) and rotational periods (0.5--5 days)
the photospheric flux is effectively veiled by accretion flux for
accretion rates \logmdot$>-9$. It is also clear from Figure
\ref{acc_dom} that the magnitude becomes brighter and the colour bluer
(in terms of optical photometry) with increasing accretion rates as
expected \citep{gullbring_1998}. The impact on the photometry of BDD
systems is discussed in the next Section for individual stars and in
Section \ref{isochrones} for populations.

@@

As shown in Figures \ref{flare_-12_183} and \ref{flare_-7_193} and
discussed in Section \ref{disc_flaring} BDD systems under vertical
hydrostatic equilibrium have highly flared discs. As discussed in
\cite{walker_2004} this increased flaring (when compared to CTTS
stars) leads to occultation at the star at lower inclination angles
and, therefore, significant changes to the SED. Increases in the
inclination angle for these systems quickly lead to a significant
proportion of the stellar flux being intercepted and reprocessed by
the highly flared disc. This reprocessing will lead to a change in the
flux levels at the shorter, bluer, wavelengths as more stellar flux is
intercepted by the disc. It will also lead to significant changes in
the flux reaching the observer from the inner and outer regions of the
disc as the system approaches edge on. Also, as discussed in Section
\ref{disc_struct} the addition of dust sublimation leads to a change
in the shape of the inner edge, where the temperature is
sufficient. Figures \ref{no_sub_183_a}, \ref{sub_179_a} and
\ref{sub_181_a} show a range in inner edge shapes, from flat walls
through concave to convex curves, caused by the radial density profile
of the disc and the dependence of the sublimation temperature on
density. This change in shape, as noted for Herbig Ae stars by
\cite{tannirkulam_2007}, will lead to changes in the characteristics
of the SED, or derived IR excess, with inclination angle
\citep{tannirkulam_2007}, for close to face on viewing angles.

@@

Figure \ref{inner_inc} shows the SEDs for the viewing angles of 0, 27,
39, and 48$^{\circ}$. The top and bottom panels of Figure
\ref{inner_inc} show SEDs for models with a curved, as exemplified by
Figure \ref{flare_-7_193}, and a flat inner wall, as shown in Figure
\ref{flare_-12_183} respectively. The specific models chosen have been
selected to minimise flux differences between the two SEDs but have
distinct inner edges. The top panel has a mass of 0.04$M_{\odot}$,
\logmdot = -8, areal coverage of 1\%, a rotational period of 5 days
and an age of 10 Myrs. The bottom panel has a the same model but with
rotation period of 0.5 days and an age of 1 Myrs. Comparing the top
and bottom panel show that flux levels for the higher accretion rate,
with the larger and curved inner edge, vary less distinctly with
inclination angle than the systems with a vertical wall. The
difference in inclination dependence of the fluxes between the curved
and vertical inner walls is marginal. This is due to differences in
the stellar flux levels and disc structures.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/inner_inc.ps}
  \caption{Figure showing SEDs of systems with a curved and flat inner
    wall as the top and bottom panels respectively. The lines show the
    SEDs at viewing angles of 0, 27, 39, and 48$^{\circ}$.
\label{inner_inc}}
\end{figure*}

@@

As stated throughout this paper the data presented are available from
a web page\footnote{http://www.astro.ex.ac.uk/research/bd\textunderscore disc}.
